---
title: "Using the conditionalPower package to estimate, at interim analysis, the probability of a successful outcome of a clinical trial"
author: "Dirk Moore"
date: "`r Sys.Date()`"

output: rmarkdown::html_vignette
#output: word_document:
#  fig_caption: true
#  toc: true
vignette: >
  %\VignetteIndexEntry{conditionalPower-vignette}
  %\VignetteEngine{knitr::rmarkdown}
#  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
 # collapse = TRUE,
  comment = ""
)
```


```{r setup, include = FALSE}
library(conditionalpower)

knitr::opts_chunk$set(
  #collapse = TRUE,
  comment = "",
  fig.width = 4,
  fig.height = 4,
	message = FALSE,
	warning = FALSE,
	tidy.opts = list(
		keep.blank.line = TRUE,
		width.cutoff = 150
		),
	options(width = 150),
	eval = TRUE
)
```

Conditional power is a tool that may used at an interim analysis  of a clinical trial to assess the likelihood of reaching a statistically significant outcome at the conclusion of the trial.

For example, suppose we are planning, at the final analysis, a one-sided 2.5 percent level test, which is powered at 85 percent. We conduct the interim analysis when 40 percent of the data (information) is available. Suppose at that point the Z-value from a test is 2.846. We can find the conditional power as follows:

```{r, echo=TRUE, eval=TRUE}
condPow2 <- condPower(tau=0.4, zz.interim=2.846, alpha=0.025, power=0.85)
```

The result is a data structure giving the conditional power and other parameters:

```{r, echo=TRUE, eval=TRUE}
condPow2
```

We may plot the predicted power, assuming that the original trend continues:

```{r, echo=TRUE, eval=TRUE, fig.show='hold', fig.width = 7, fig.height =5}
plotCondPower(condPow2)
```


